---
title: "Dimension Reduction"
author: "Prof SF Koch"
format: revealjs
editor: visual
---

# Dimension Reduction


## PCA background

- Principle Components Analysis
    + Find linear combination of something
    + The linear combination should maximize the variance of the combination
    
- Why Use PCA?
    + Imagine a bunch of wines on the table (all different)
    + They have different characteristics: nose, colour, strength, tanins...
    + Some of those characteristics might measure related properties
    + Use PCA to find "fewer" characteristics

## PCA intuition

+ Does not discard information
+ Re-imagines information as linear combination
+ We are going to transform the data to be orthogonal
    + That is why we need Eigenvalues
    + That is why we need data that is only 'scaled'
+ Picking out new characteristics that are truly different
+ But, difficult to "interpret" in use
  
## Basic PCA: Part 1

+ Assume an $n times k$ matrix of mean-centered data $X$
+ We want to find weights $w \in [0,1]$ that yield PCs from the data
+ Where $T$ are the PCs

$$
T = XW
$$

## Basic PCA: Part 2

+ We want to find weights that maximise the variance of the data
+ Thus, we want to solve, something like this, here is for the first weight vector

$$ 
w_1 = \arg \max_{||w|| = 1} w'X'Xw
$$

+ The solution yields $w$ to be an eigenvector

## PCA process: steps 1 - 2

1. Standardize the data
    + $s_{x_i}$ is the standard deviation of that variable
    + There are times, when we use a different standardization

$$ 
\tilde x_i = \frac{x_i - \overline{x}}{s_{x_i}}
$$

2. Compute the variance-covariance matrix for $\tilde{X}$

$$
V(\tilde X,\tilde X) = \tilde X'\tilde X
$$

## PCA prcess: steps 3 - 5

3. Compute the eigenvectors and values of $C$ - see previous slides

4. Create feature vector $F$: eigenvectors we want to keep

5. Recast the data

$$
\tilde X_{\text{recast}} =F' \tilde X
$$

# Example with Income and Expenditure

```{r simplecor, dependson="lcs2"}

cor.exinc <- cor(lasso.df$y.ihs,lasso.df$x.ihs)
```


## Recalling from above

+ Income/expenditure was ridged or lassoed
+ Thus, the dimensions were probably similar
+ Let us check the correlation 
    + Should have done earlier, of course
    + Correlation is: ```r round(cor.exinc,digits = 3)```
+ So, lets "combine"

## Eigenvalues

```{r pca1, dependson="lcs2"}

xy.df <- lasso.df %>% dplyr::select(y.ihs,x.ihs) 

pr.inc.exp <- prcomp(xy.df, scale = TRUE)

pr.var <- pr.inc.exp$sdev^2
pve <- pr.var/sum(pr.var)

```

+ Let us find the two pieces
    + Vector 1  = [ ```r pr.inc.exp$rotation[,1]``` ]
    + Vector 2 = [ ```r pr.inc.exp$rotation[,2]``` ]

+ Rather similar
    + Not surprising in this setting
    + Only two dimensions
    
+ With more dimensions, we have more than quadratic vector

+ Share of variation in first component = ```r pve[1]```

## A plot of the data with eigenvectors

```{r pca-data, dependson="pca1", fig.cap = "Plot of Income and Expenditure"}

plot(x=xy.df$x.ihs, y=xy.df$y.ihs, type="p", 
  xlab = "Expenditure IHS", ylab = "Income IHS")
arrows(mean(xy.df$x.ihs), mean(xy.df$y.ihs),
       pr.inc.exp$rotation[2,1] + mean(xy.df$x.ihs),
       pr.inc.exp$rotation[1,1] + mean(xy.df$y.ihs),
       col = "red", lwd = 3)
arrows(mean(xy.df$x.ihs), mean(xy.df$y.ihs),
       pr.inc.exp$rotation[2,2] + mean(xy.df$x.ihs),
       pr.inc.exp$rotation[1,2] + mean(xy.df$y.ihs),
       col = "blue", lwd = 3)

```

    
## A plot of the loadings

```{r pca2, dependson="pca1", fig.cap = "Plot of Loadings"}

plot(prcomp(xy.df, scale. = TRUE),
     main = "Loadings plot",
     xlab = "Component dimensions")
```


## Regression with first component: estimates and prediction error

```{r revised-regression, dependson="ridge.result; cv-ridge; pca1", results="asis"}


train.rows <- sample(nrow(df.x),3*nrow(df.x)/4)

xy1 <- t(as.matrix(pr.inc.exp$rotation[,1])) %*% t(as.matrix(xy.df)) 
lasso.df2 <- lasso.df %>% add_column(xy.ihs = as.vector(xy1))

df.train <- lasso.df2[train.rows,]
df.test <- lasso.df2[-train.rows,]

reg.cv <- lm(min.ihs ~ xy.ihs + kids + adults + age + male 
                     + province + pop.group, 
             data=df.train)

reg.test.pred2 <- predict(reg.cv, 
                         newdata = df.test)

xtable(summary(reg.cv)$coefficients[2:5,])


```

+ Prediction error from updated model = ```r round(mean((reg.test.pred2 - df.test$min.ihs)^2),4)```

# Asset Index : Commonly Used in Applied Micro

## Living standards

- Measure living standards
    + Could use income/expenditure
    + Could use per capita income/expenditure
    + Could use per adult equivalent income/expenditure
- Single dimension measures

## Multi-dimensional living standards

- Obvious way to order/rank in single dimensions
- What if combining dimensions?
- May not be obviously ordered [@wittenbergleibbrandt2017]
- So, find a way to combine


## Proposed ideals - to be a "living standard" 

- *Principle 1* Monotonicity
    + With more of something, higher index value
    + Intuitive

- *Principle 2* Absolute zero
    + Having nothing is the bottom
    + Needed for reference
    + Referred to as 'ratio-scale'
    + Basically, no negatives
    + Requires us to limit attention to **goods**

- *Principle 3* Robustness
    + Type of variable should not matter
    + Categorical or continuous

## Some Basic Points

- PCA is a procedure
    + It is devoid of a view
    + It is also "unsupervised"
    + Sometimes that view might matter
    + True for all procedures!!!!!

- Do not apply, just to apply
    + Think about the inputs
    + Requires thought about meaning
    + How is it being used?
    + For what is it being used?
    
## An Alternative

1. Choose variables for asset index
2. Run PCA
3. Check PCA weights
    + If all positive, ok, you are done
    + If some negative, reconsider
4. Run uncentered PCA
    + Do not standardize variables as in step 1
    + Instead, standardize through division by  means
    + Guaranteed positive weights
    + Otherwise, same steps
5. Also, Re-run PCA dropping negative weight variables
6. Check
    + Look at rankings across indexes
    + Look at inequality across indexes
    + Important enough to matter?

# Ridge and Lasso

## Opening queries   

1. What does least squares do?
1. Optimum based on RSS
1. Will it give us zeroes?
    a. Statistically, yes
    a. Should we ignore them?

## Related queries

1. How much data is there in a StatsSA survey?
    a. ±20 000 households
    a. ±80 000 individuals
    a. ±3 different data sets
    a. trillions and trillions of relationships?
1. What variables should be included?
    a. Sure, we use theory to limit
    a. But, data is usually less precise than theory
    
## Attempt to simplify

1. Dimension Reduction
    a. Human 'interest' greatly limits regression controls
        i. No one wants a 50 page table of results
        i. No one wants to read the description of that table
    a. $\ell_1$ shrinkage
    a. $\ell_2$ shrinkage
    a. One can also consider testing groups of coefficients and dropping
1. Dimension Redefinition 
    a. Principal components (centred or uncentered)
    a. Multiple correspondence

## Basic OLS

1. As we know

$$
\min_{\beta} \sum_{i=1}^n \left(  y_i - \beta x_i \right)^2
$$

2. Noise will likely never give us an actual zero
3. But, we are likely to get statistical insignificance
4. There might also be collinearity 'fears'


## Illustration

```{r sillyplot, echo=F, fig.cap="True and estimated 0: A bit contrived"}

e <- rnorm(100,0,1)
y <- 0.5 + e
x1 <- runif(100,0,5)
x2 <- x1/1000
model1 <- lm(y~x1)

plot(x1,y)
abline(h=mean(y),col="black",lwd=2)
abline(reg=model1,col="red",lwd=2,lty=2)
legend("topleft",
       lwd=c(2,2),
       lty=c(1,2),
       c("True 0","Fitted 0"))

```

## Ridging

1. We now add a penalty to the model


$$
\min_{\beta} \sum_{i=1}^n \left(  y_i - \beta x_i \right)^2 + \lambda \sum_{j=1}^p \beta_j^2  
$$

2. This is an $\ell_2$ penalty
    a. It is squared
    a. It is smooth
    a. So, 0 is not so likely
    a. All parameters "pushed down"


## OLS estimates

```{r ols-tab, dependson="lcs2", results="asis", size = "tiny"}

linear.model <- lm(min.ihs ~ x.ihs + y.ihs + kids + adults + age + male + province + pop.group,
                   data = lasso.df)

linear.summary <- summary(linear.model)$coefficients[2:6,]
xtable(linear.summary)
```

+ Quick discussion
+ Next will do Ridge and compare to these


```{r ridge, dependson="data", results='asis', size = "tiny"}

# this does not take typical formula structure


mnihs <- lasso.df$min.ihs

ridge.reg = glmnet(df.x, mnihs, 
                   nlambda = 100, #default
                   alpha = 0, #ridge
                   family = 'gaussian')


```

## Compare Some Results 

```{r ridge.result, dependson="ridge", results="asis"}

# ignoring the intercept in this
res1 <- coef(ridge.reg)[2:6,1]
res2 <- coef(ridge.reg)[2:6,50]
res3 <- coef(ridge.reg)[2:6,100]
olscoef <- coef(linear.model)[2:6]
lambda1 <- paste0("lambda=",round(ridge.reg$lambda[1],3))
lambda2 <- paste0("lambda=",round(ridge.reg$lambda[50]),3)
lambda3 <- paste0("lambda=",round(ridge.reg$lambda[100],3))
comparison <- cbind(olscoef,res1,res2,res3)
colnames(comparison) <- c("OLS",lambda1,lambda2,lambda3)
xtable(comparison,digits=4)

```

## Lambda plot - quite a few coefficients here...

```{r, dependson="ridge", fig.cap="Lambda plot"}

plot(ridge.reg, xvar = "lambda")

```

## Cross-Validated Plot

```{r cv-ridge, dependson="ridge", fig.cap="CV Plot"}

train.rows <- sample(nrow(df.x),3*nrow(df.x)/4)
df.x.train <- df.x[train.rows,]
df.x.test <- df.x[-train.rows,]
mnihs.train <- mnihs[train.rows] # this is only one dimension, so no "," at the end
mnihs.test <- mnihs[-train.rows]

cv.ridge <- cv.glmnet(df.x.train,mnihs.train, alpha = 0)
plot(cv.ridge)

```


## Cross-Validated Lambda and MSE

```{r cv-ridge-lam, dependson="cv-ridge", results="asis"}

lambda.ridge <- cv.ridge$lambda.min
paste("Optimal value (Ridge) for lambda = ", round(lambda.ridge,4)," ")

print(" ")

ridge.test.prediction <- predict(cv.ridge,s=lambda.ridge,newx = df.x.test)
paste("Prediction error from optimal Ridge model =", 
      round(mean((ridge.test.prediction - mnihs.test)^2),4))
```

## Refit with Optimal Test Lambda

```{r, finalestimates, dependson="cv-ridge-lambda;ridge; ridge.result", results="asis"}

ridge.optimal <- glmnet(df.x, mnihs, alpha = 0)
ridge.opt.predict <- predict(ridge.optimal,type="coefficients",s = lambda.ridge)[2:6,]

results.compare <- cbind(olscoef,ridge.opt.predict)
colnames(results.compare) <- c("OLS","Optimal Ridge")
xtable(results.compare,digits=5)
```

+ A bit of a difference?
+ Possible to use in probit [@lockingetal]
+ And, many other models

# LASSO

## An Adjustment to the Ridge

1. We revise the penalty to the model


$$
\min_{\beta} \sum_{i=1}^n \left(  y_i - \beta x_i \right)^2 + \lambda \sum_{j=1}^p \left| \beta_j \right|
$$

2. This is an $\ell_1$ penalty
    a. It is the absolute value
    a. It is "pointy", like a check function
    a. So, 0 is more likely
    
## Some Coefficients are Zeroed

```{r lasso, dependson="cv-ridge", fig.cap="Lasso Coefficients"}

lasso.reg = glmnet(df.x.train, mnihs.train, nlambda = 100, alpha = 1)
plot(lasso.reg)

```


## Cross-Validated Plot

```{r cv-lasso, dependson="cv-ridge", fig.cap="CV Plot"}

cv.lasso <- cv.glmnet(df.x.train, mnihs.train, alpha = 1)
plot(cv.lasso)

```


## Cross-Validated Lambda and MSE

```{r cv-lasso-lam, dependson="cv-lasso", results="asis"}

lambda.lasso <- cv.lasso$lambda.min

paste("Optimal value for lambda (LASSO) = ", round(lambda.lasso,4)," ")

print(" ")

lasso.test.prediction <- predict(cv.lasso,s=lambda.lasso,newx = df.x.test)

paste("Prediction error from optimal LASSO model =", 
      round(mean((lasso.test.prediction - mnihs.test)^2),4))

```

## Refit with Optimal Test Lambda

```{r, finalecompare, dependson="cv-lasso-lam;cv-ridge-lambda;ridge.result", results="asis"}

lasso.optimal <- glmnet(df.x, mnihs, alpha = 1)
lasso.opt.predict <- predict(lasso.optimal,type="coefficients",s = lambda.lasso)[2:6,]

results.compare <- cbind(olscoef,lasso.opt.predict,ridge.opt.predict)
colnames(results.compare) <- c("OLS","Optimal Lasso","Optimal Ridge")
xtable(results.compare,digits=5)
```

+ Hmmm, interesting difference?


## OLS CV Value for Comparison

```{r, dependson="ridge.result; cv-ridge", results="asis"}
df.train <- lasso.df[train.rows,]
df.test <- lasso.df[-train.rows,]
reg.cv <- lm(min.ihs ~ x.ihs + y.ihs + kids + adults + age + male 
                     + province + pop.group, 
             data=df.train)

reg.test.pred <- predict(reg.cv, 
                         newdata = df.test)

paste("Prediction error from OLS model =", 
      round(mean((reg.test.pred - df.test$min.ihs)^2),4))

```


+ As suggested, OLS can be better
+ But, it is really really close
+ Ridge "wins" by a bit

# Principal Components Analysis

# Eigenvectors and Eigenvalues

## Eigenvector or characteristic value

+ Vector whose direction is unchanged if transformed
+ Define as $v$ with linear transformation $T$
+ And, assume a constant scale factor $\lambda$

$$
Tv = \lambda v
$$

+ $\lambda$ is the characteristic root or eigenvalue

## Some generic concepts

+ Vector has a magnitude and direction
+ Linear transforms rotate, stretch and or shear the vector
+ Eigenvectors only stretch...
+ So, return to above
    + Consider matrix $A$ and vector $v$
    + If $Av$ only scales $v$ by a factor $\lambda$...
    + $v$ is an eigenvector of $A$, and $Av=\lambda v$
+ Thus, our interest is in solving for $v$

$$
(A - \lambda I) v = 0
$$

+ This only "works" if - we are not inverting...

$$
| A - \lambda I | =  0
$$

## Simplistic example: Part 1

+ Take a simple $A$

$$
A = \begin{bmatrix} 2 & 1 \\ 1 & 2 \end{bmatrix}
$$

+ Write down the determinant and solve for $\lambda$

$$
\begin{split}
|A - \lambda I | &= \begin{bmatrix} 2-\lambda & 1 \\ 1 & 2-\lambda \end{bmatrix} \\
&= (2-\lambda)^2 - 1 = \lambda^2 - 4\lambda + 3 \\
\lambda &= \{1,3\}
\end{split}
$$

## Simplistic example: Part 2

+ Solve for $v$, use $\lambda = 1$

$$
\begin{split}
(A - \lambda I ) v &= \begin{bmatrix} 2-1 & 1 \\ 1 & 2-1 \end{bmatrix}  
   \begin{bmatrix} v_1 \\ v_2 \end{bmatrix} = \begin{bmatrix} 0 \\ 0 \end{bmatrix} \\
   &= \begin{bmatrix} 1 & 1 \\ 1 & 1 \end{bmatrix} \begin{bmatrix} v_1 \\ v_2 \end{bmatrix} = 
     \begin{bmatrix} 0 \\ 0 \end{bmatrix}  \\
&= \begin{bmatrix} v_1 + v_2 \\ v_1 + v_2 \end{bmatrix}  = 
     \begin{bmatrix} 0 \\ 0 \end{bmatrix} \\
v_1 &= -v_2 \\
v = \begin{bmatrix} 1 \\ -1 \end{bmatrix} \\
\end{split}
$$


+ Do the same to find for $\lambda = 3$


