---
title: "Regression Analysis"
subtitle: "Theory, Review and Preliminary Ideas"
author: "Steve Koch"
date: "today"
format: 
  revealjs:
    theme: solarized
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
editor: source
bibliography: 
  - module.bib
  - software-refs.bib
  - data-slides.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,
                      echo=FALSE,
                      autodep=TRUE,
                      message=FALSE,
                      warning=FALSE,
                      dev='pdf',
                      out.width="90%",
                      fig.asp=0.65,
                      fig.align='center')
knitr::opts_knit$set(kable.force.latex = TRUE)


library(Matching) 
library(ggplot2)
library(xtable)
set.seed(42)
library(sandwich)
library(lmtest)
library(car)
library(nlstools)
library(MASS)
library(nnet)
library(tidyverse)
library(arm)
library(mvtnorm)
library(lme4)
library(multiwayvcov)
library(clusterSEs)
library(haven)
```


# Expectations

## Finite population mean

- Finite (implicitly discrete) 
    + $E[X] = \sum_{x\in X} x \cdot p(x)$
    + $x$ is a realisation in the population of $X$
    + $p$ is the probability density
    + Classic q: What is it for the roll of a die?

## Infinite population mean

- Infinite 
    + $E[X] = \int_x x \cdot f(x)dx$
    + $x$ is continuous and lies in the population of $X$
    + $f$ is the probability density function
    + Classic q: What is it for a uniform distribution on $[a,b]$?

- Note: $x$ could be a function, too.

## Expectations rules 

1. $E[X \pm k] = E[X] \pm k$
1. $E[k \cdot X] = k \cdot E[X]$
1. $E[X \pm Y] = E[X] \pm E[Y]$
1. $E[X \cdot Y] = E[X] \cdot E[Y]$
    a. Only if $X$ and $Y$ are independent
    a. What is independence? 
    
## Independence reviewed

- Set definition

$$
P(A \cap B) = P(A)P(B)
$$
 
- Conditional set definition

$$
P(A|B) = \frac{P(A \cap B)}{P(B)} = \frac{P(A)P(B)}{P(B)} = P(A)
$$

## In words

- The "occurrence" of B does not affect the "occurrence" of A
- $A$ and $B$ can still have common elements
- Independence is not mutually exclusive

$$
P(A \cap B) = \emptyset
$$


# Variances
    
## Population measures

- $V[X] = E\left[ (X-E[X])^2 \right]$
- $V[X] = E[X^2] - E[X]^2 \neq 0$
- Calculate variance from a die
- $C[X,Y] = E[(X-E[X])(Y-E[Y])]$
    
## Variance rules

1. $V[X \pm k] = V[X]$
1. $V[k \cdot X] = k^2 \cdot V[X]$
1. $V[X \pm Y] = V[X] \pm V[Y]$ if independent
1. $V[X \pm Y] = V[X] + V[Y] \pm C[X,Y]$


# Conditional Expectations

## Definitions and Theory

- Assuming $B$ is a non-zero probability event

$$
P(A|B) = \frac{P(A \cap B)}{P(B)} 
$$

- A conditional density

$$
p(x|B) = P(X=x|B) = \frac{P(X=x,B)}{P(B)}
$$

## Underpinning regression

- A conditional expectation

$$
E[X|B] = \sum_{x \in X} x \cdot p(x|B)
$$


## More common discrete

- Two discrete random variables $X$, $Y$, with conditional joint density 

$$
f_{Y|X} (y|x) = \frac{P(Y=y,X=x)}{P(X=x)} = \frac{f_{Y,X} (y,x)}{f_X(X)}
$$

- Conditional expectation

$$
E[Y|X=x] = \sum_{y \in Y} y \cdot f_{Y|X}(y|x)
$$

## Separate this

- Overall (partition theorem)

$$ 
E[Y] = \sum_{x \in X} E[Y|X=x] \cdot P(X=x)
$$

## Continuous conditional expectation version

- Conditional expectation

$$
E[Y|X=x] = \int_{-\infty}^{\infty} y \cdot f_{Y|X}(y|x)dy
$$

- Overall (partition theorem)

$$
E[Y] = \int_{-\infty}^{\infty} E[Y|X=x] \cdot f_X(x)dx
$$

# Linear Projections


## Property I

- Let $(Y, X)$ have some joint distribution with finite variance
- The linear projection of $Y$ onto $X$ is $\hat{Y} = a + b X$

$$
\begin{aligned}
a =& E[Y] - bE[X] \\
b =& \frac{C[X,Y]}{V[X]} 
\end{aligned}
$$


## Property II

- Define $Y^* = Y - (a + bX)$
- What is $E[Y^*]$?
- What is $C[Y^*,X]$?
- Take out paper and try...
- On the next few slides are some useful practice questions...

## Example 1

<!--
\begin{table}
\begin{tabular}{l|c|c|}
$\downarrow D$, $X \rightarrow$ & 0 & 1 \\
\hline
 0 & 0.1 & 0.2 \\
 1 & 0.5 & 0.2 \\
\end{tabular}
\end{table}
-->

| ↓ D \ X → |  0  |  1  |
|:----------|:---:|:---:|
| 0         | 0.1 | 0.2 |
| 1         | 0.5 | 0.2 |

- Use R to compare to "paper" analysis
- Find linear projection of $D$ on $X$
- Find $C[D^*,X]$
- Are $D^*$ and $X$ independent?

## Example 2

<!--
\begin{table}
\begin{tabular}{l|c|c|}
$\downarrow X$, $Y \rightarrow$ & 0 & 1 \\
\hline
 0 & 0.2 & 0.1 \\
 1 & 0.1 & 0.3 \\
 2 & 0.2 & 0.1 \\
\end{tabular}
\end{table}
-->

| ↓ X \ Y → |  0  |  1  |
|:----------|:---:|:---:|
| 0         | 0.2 | 0.1 |
| 1         | 0.1 | 0.3 |
| 2         | 0.2 | 0.1 |

- Use R to compare to "paper" analysis
- Find the linear projection of $Y$ onto $X$
- Find $E[Y|X=\{1,2,3\}]$
- Is $\hat{Y}$ the same as $E[Y|X]$?



## Conditional expectations summary
    
+ The overall mean can be partitioned
+ The conditional mean and the unconditional mean (binary case) are the same only if the conditioning variable is independent

## Projection summary

+ A linear projection may not match the conditional expectation
+ The latter may be nonlinear
+ A residual may not be independent

# Basic Model

## Generic Model 

- Linear projection with error

$$
y = X\beta + u
$$

- The solution concept

$$
\hat{\beta} = \underset{b}{\operatorname{argmin}} \text{ }  E
\left[ \left( y - X \beta \right)' \left( y - X \beta \right) \right] 
$$

## Generic Model FOC 


$$
\begin{aligned}
2 [X' y - X' X \hat{\beta}] =& 0 \\
X' X \hat{\beta} =&X' y \\
[X'X]^{-1} X'X  \hat{\beta} =& [X'X]^{-1} X'y  \\
\hat{\beta} =& [X'X]^{-1}X'y 
\end{aligned}
$$

## The Sample Analog

$$
\begin{aligned}
E[X'X]^{-1} \rightarrow & \left[ \frac{X'X}{n} \right]^{-1} = n \left[
  \sum_{j=1}^n X_j X_j' \right]^{-1}\\
E[X'y] \rightarrow & \frac{X'y}{n} =  \frac{1}{n}\sum_{j=1}^n X_j y_j
\end{aligned}
$$

## Rewrite the solution 

+ Ignoring conditioning on $X$ for simple notation

$$
\begin{aligned}
\hat{\beta} =& [X'X]^{-1}[X' (X\beta+u)] \\
  =& [X'X]^{-1}[X' X\beta]+[X'X]^{-1}X'u \\
  =& \beta + [X'X]^{-1}X'u
  \end{aligned}
$$

## Check the expectation (with respect to $X$)

$$
\begin{aligned}
E \left[ \hat{\beta} | X \right] =& \beta + E \left[ [X'X]^{-1}X'u | X
\right] \\
 =& \beta +  [X'X]^{-1} E[X'u | X] 
 \end{aligned}
$$

- Is it unbiased?
- How? Why?

## Variance of Estimate

- The definition

$$
V(\hat{\beta}|X) = E \left[ \left( \hat{\beta} - \beta \right) \left( \hat{\beta} - \beta
  \right)' \bigg| X \right]
$$

- Substitution and rearrangement

$$
V(\hat{\beta}| X) = E \Biggl[ [X'X]^{-1}X'uu'X[X'X]^{-1} \Bigg|X \Biggr] 
$$

## Heteroskedasticity

- The above implies/allows heteroskedasticity
- Rewrite the variance 
- Use at observation level to avoid matrix notation.

$$
\begin{aligned}
E \left[ (y_i - X_i'\beta)^2 | X \right] =&  E \left[ \left( (y_i - E[Y_i|X_i])
  + (E[y_i|X_i] -X_i'\beta) \right)^2 \Big|X_i \right] \\
  =& V[y_i|X_i] + (E[y_i|X_i] -X_i'\beta)^2
  \end{aligned}
$$

## Rewrite and rearrange

- Simple add and subtract of $E[y_i|X_i]$
- Run the expectation through
- Even if the variance of $y$ is constant, conditional on $X$, the linear approximation of a nonlinear CEF will result in heteroscedasticity

# Clustering

```{r}
#| label: clusterdata
#| echo: false

#- Analysis of Clustered Data
#- Courtesy of Dr. Yuki Yanai, 
#- http://yukiyanai.github.io/teaching/rm1/contents/R/clustered-data-analysis.html

gen_cluster <- function(param = c(.1, .5), n = 1000, n_cluster = 50, rho = .5) {
  # Function to generate clustered data
  # Required package: mvtnorm
  
  # individual level
  Sigma_i <- matrix(c(1, 0, 0, 1 - rho), ncol = 2)
  values_i <- rmvnorm(n = n, sigma = Sigma_i)
  
  # cluster level
  cluster_name <- rep(1:n_cluster, each = n / n_cluster)
  Sigma_cl <- matrix(c(1, 0, 0, rho), ncol = 2)
  values_cl <- rmvnorm(n = n_cluster, sigma = Sigma_cl)
  
  # predictor var consists of individual- and cluster-level components
  x <- values_i[ , 1] + rep(values_cl[ , 1], each = n / n_cluster)
  
  # error consists of individual- and cluster-level components
  error <- values_i[ , 2] + rep(values_cl[ , 2], each = n / n_cluster)
  
  # data generating process
  y <- param[1] + param[2]*x + error
  
  df <- data.frame(x, y, cluster = cluster_name)
  return(df)
}

# Simulate a dataset with clusters and fit OLS
# Calculate cluster-robust SE when cluster_robust = TRUE
cluster_sim <- function(param = c(.1, .5), n = 1000, n_cluster = 50,
                        rho = .5, cluster_robust = FALSE) {
  # Required packages: mvtnorm, multiwayvcov
  df <- gen_cluster(param = param, n = n , n_cluster = n_cluster, rho = rho)
  fit <- lm(y ~ x, data = df)
  b1 <- coef(fit)[2]
  if (!cluster_robust) {
    Sigma <- vcov(fit)
    se <- sqrt(diag(Sigma)[2])
    b1_ci95 <- confint(fit)[2, ]
  } else { # cluster-robust SE
    Sigma <- cluster.vcov(fit, ~ cluster)
    se <- sqrt(diag(Sigma)[2])
    t_critical <- qt(.025, df = n - 2, lower.tail = FALSE)
    lower <- b1 - t_critical*se
    upper <- b1 + t_critical*se
    b1_ci95 <- c(lower, upper)
  }
  return(c(b1, se, b1_ci95))
}

# Function to iterate the simulation. A data frame is returned.
run_cluster_sim <- function(n_sims = 1000, param = c(.1, .5), n = 1000,
                            n_cluster = 50, rho = .5, cluster_robust = FALSE) {
  # Required packages: mvtnorm, multiwayvcov, dplyr
  df <- replicate(n_sims, cluster_sim(param = param, n = n, rho = rho,
                                      n_cluster = n_cluster,
                                      cluster_robust = cluster_robust))
  df <- as.data.frame(t(df))
  names(df) <- c('b1', 'se_b1', 'ci95_lower', 'ci95_upper')
  df <- df |> 
    mutate(id = 1:n(),
           param_caught = ci95_lower <= param[2] & ci95_upper >= param[2])
  return(df)
}

```

## Generate some data

```{r}
#| eval: false

#- Analysis of Clustered Data
#- Courtesy of Dr. Yuki Yanai, 
#- http://yukiyanai.github.io/teaching/rm1/contents/R/clustered-data-analysis.html

gen_cluster <- function(param = c(.1, .5), n = 1000, n_cluster = 50, rho = .5) {
  # Function to generate clustered data
  # Required package: mvtnorm
  
  # individual level
  Sigma_i <- matrix(c(1, 0, 0, 1 - rho), ncol = 2)
  values_i <- rmvnorm(n = n, sigma = Sigma_i)
  
  # cluster level
  cluster_name <- rep(1:n_cluster, each = n / n_cluster)
  Sigma_cl <- matrix(c(1, 0, 0, rho), ncol = 2)
  values_cl <- rmvnorm(n = n_cluster, sigma = Sigma_cl)
  
  # predictor var consists of individual- and cluster-level components
  x <- values_i[ , 1] + rep(values_cl[ , 1], each = n / n_cluster)
  
  # error consists of individual- and cluster-level components
  error <- values_i[ , 2] + rep(values_cl[ , 2], each = n / n_cluster)
  
  # data generating process
  y <- param[1] + param[2]*x + error
  
  df <- data.frame(x, y, cluster = cluster_name)
  return(df)
}

# Simulate a dataset with clusters and fit OLS
# Calculate cluster-robust SE when cluster_robust = TRUE
cluster_sim <- function(param = c(.1, .5), n = 1000, n_cluster = 50,
                        rho = .5, cluster_robust = FALSE) {
  # Required packages: mvtnorm, multiwayvcov
  df <- gen_cluster(param = param, n = n , n_cluster = n_cluster, rho = rho)
  fit <- lm(y ~ x, data = df)
  b1 <- coef(fit)[2]
  if (!cluster_robust) {
    Sigma <- vcov(fit)
    se <- sqrt(diag(Sigma)[2])
    b1_ci95 <- confint(fit)[2, ]
  } else { # cluster-robust SE
    Sigma <- cluster.vcov(fit, ~ cluster)
    se <- sqrt(diag(Sigma)[2])
    t_critical <- qt(.025, df = n - 2, lower.tail = FALSE)
    lower <- b1 - t_critical*se
    upper <- b1 + t_critical*se
    b1_ci95 <- c(lower, upper)
  }
  return(c(b1, se, b1_ci95))
}

# Function to iterate the simulation. A data frame is returned.
run_cluster_sim <- function(n_sims = 1000, param = c(.1, .5), n = 1000,
                            n_cluster = 50, rho = .5, cluster_robust = FALSE) {
  # Required packages: mvtnorm, multiwayvcov, dplyr
  df <- replicate(n_sims, cluster_sim(param = param, n = n, rho = rho,
                                      n_cluster = n_cluster,
                                      cluster_robust = cluster_robust))
  df <- as.data.frame(t(df))
  names(df) <- c('b1', 'se_b1', 'ci95_lower', 'ci95_upper')
  df <- df |> 
    mutate(id = 1:n(),
           param_caught = ci95_lower <= param[2] & ci95_upper >= param[2])
  return(df)
}

```


## Data with no clusters

```{r}
#| label: noclustBeta
#| dependson: clusterdata
#| echo: false
#| fig-cap: "Histogram of estimates"

# Distribution of the estimator and confidence intervals
sim_params <- c(.4, 0)   # beta1 = 0: no effect of x on y
sim_nocluster <- run_cluster_sim(n_sims = 10000, param = sim_params, rho = 0)
hist_nocluster <- ggplot(sim_nocluster, aes(b1)) +
  geom_histogram(color = 'black') +
  geom_vline(xintercept = sim_params[2], color = 'red')
print(hist_nocluster)
```

## No clustering power

```{r}
#| label: noclustSE
#| dependson: clusterdata
#| echo: false
#| fig-cap: "Statistical significance of estimates"

ci95_nocluster <- ggplot(sample_n(sim_nocluster, 100),
                         aes(x = reorder(id, b1), y = b1, 
                             ymin = ci95_lower, ymax = ci95_upper,
                             color = param_caught)) +
  geom_hline(yintercept = sim_params[2], linetype = 'dashed') +
  geom_pointrange() +
  labs(x = 'sim ID', y = 'b1', title = 'Randomly Chosen 100 95% CIs') +
  scale_color_discrete(name = 'True param value', labels = c('missed', 'hit')) +
  coord_flip()
print(ci95_nocluster)

sim_nocluster_mean <- sim_nocluster |> 
  summarize(type1_error = 1 - sum(param_caught)/n()) |> 
  pull()

```

## Plot code (with no clustering in data)


```{r}
#| eval: false

ci95_nocluster <- ggplot(sample_n(sim_nocluster, 100),
                         aes(x = reorder(id, b1), y = b1, 
                             ymin = ci95_lower, ymax = ci95_upper,
                             color = param_caught)) +
  geom_hline(yintercept = sim_params[2], linetype = 'dashed') +
  geom_pointrange() +
  labs(x = 'sim ID', y = 'b1', title = 'Randomly Chosen 100 95% CIs') +
  scale_color_discrete(name = 'True param value', labels = c('missed', 'hit')) +
  coord_flip()
print(ci95_nocluster)

sim_nocluster_mean <- sim_nocluster |> 
  summarize(type1_error = 1 - sum(param_caught)/n()) |> 
  pull()

```


## Data with clusters histogram

```{r}
#| label: clusterBeta
#| dependson: clusterdata
#| echo: false
#| fig-cap: "Histogram of estimates"

#Data with clusters
sim_params <- c(.4, 0)   # beta1 = 0: no effect of x on y
sim_cluster_ols <- run_cluster_sim(n_sims = 10000, param = sim_params)
hist_cluster_ols <- hist_nocluster %+% sim_cluster_ols
print(hist_cluster_ols)

```

## Uncorrected clusters statistical power

```{r}
#| label: clusterSE
#| dependson: clusterdata
#| echo: false
#| fig-cap: "Statistical significance of estimates"

#Confidence interval
ci95_cluster_ols <- ci95_nocluster %+% sample_n(sim_cluster_ols, 100)
print(ci95_cluster_ols)

sim_cluster_ols_mean <- sim_cluster_ols |> 
  summarize(type1_error = 1 - sum(param_caught)/n()) |> 
  pull()
```

## Corrected clusters statistical power

```{r}
#| label: clusterSEfixed
#| dependson: clusterdata
#| fig-cap: "Statistical significance of estimates"
#| echo: false

#clustered robust
sim_params <- c(.4, 0)   # beta1 = 0: no effect of x on y
sim_cluster_robust <- run_cluster_sim(n_sims = 10000, param = sim_params,
                                      cluster_robust = TRUE)

#hist_cluster_robust <- hist_nocluster %+% sim_cluster_ols
#print(hist_cluster_robust)

#Confidence Intervals
ci95_cluster_robust <- ci95_nocluster %+% sample_n(sim_cluster_robust, 100)
print(ci95_cluster_robust)

sim_cluster_robust_mean <- sim_cluster_robust |> 
  summarize(type1_error = 1 - sum(param_caught)/n()) |> 
  pull()
```

## Correcting for clusters code

```{r}
#| eval: false

# code from Cunningham

#clustered robust
sim_params <- c(.4, 0)   # beta1 = 0: no effect of x on y
sim_cluster_robust <- run_cluster_sim(n_sims = 10000, param = sim_params,
                                      cluster_robust = TRUE)

#hist_cluster_robust <- hist_nocluster %+% sim_cluster_ols
#print(hist_cluster_robust)

#Confidence Intervals
ci95_cluster_robust <- ci95_nocluster %+% sample_n(sim_cluster_robust, 100)
print(ci95_cluster_robust)

sim_cluster_robust_mean <- sim_cluster_robust |> 
  summarize(type1_error = 1 - sum(param_caught)/n()) |> 
  pull()
```


## Quick review of statistical power

1. No clustering, no correction "False rejection rate" = `r round(sim_nocluster_mean,3)`

1. Clustering, no correction "False rejection rate" = `r round(sim_cluster_ols_mean,3)`

1. Clustering, with correction "False rejection rate" = `r round(sim_cluster_robust_mean,3)`

# Further Regression Features

## Mechanical OLS and projection

+ What if we consider nonlinear conditional expectations in linear models?
+ We will apply the matrix math to see the solution
+ It is good practice and you should see if you can replicate the result

## Look at CEF

+ We explore three important properties
+ These properties offer deep insight into OLS
+ We also look into Frisch-Waugh-Lovell theorem
+ I recommend trying to 'replicate' the "proofs"/properties


## Discrete Value Covariates

- Suppose $X = \{1, 2, \dots, K \}$
- Define $D_k = 1$ if $X=k$
- Define the (nonlinear?) conditional expectation

$$
E[y|X=x] = \begin{cases}
  \beta_1 & \text{ if } X = 1 \\
  \beta_2 & \text{ if } X = 2 \\
  \vdots \\
  \beta_K & \text{ if } X = K
  \end{cases} = \sum_k D_k \beta_k 
$$

## Linear in parameters regression

- Saturated regression

$$
y_i = \sum_k \beta_k D_{ik} + \varepsilon_i
$$

- For K = $\{1, 2 \}$ and $n = 3$

$$
X = \begin{bmatrix} 1 & 0 \\
                    0 & 1 \\
                    0 & 1 
                    \end{bmatrix}
$$

## Matrix analysis 

$$
X'X = \begin{bmatrix} 1 & 0 \\
                      0 & 2
                      \end{bmatrix}
$$

```{r}
x <- matrix(c(1,0,0,0,1,1),ncol = 2)
xtx <- t(x) %*% x
xtx.inv <- solve(xtx)
```


$$
[X'X]^{-1} = \begin{bmatrix} 1 & 0 \\
                      0 & 0.5
                      \end{bmatrix}
$$

## Saturated Regression $X'Y$

- Define $y$

$$ 
y' = \begin{bmatrix} y_1 & y_2 & y_3 \end{bmatrix}
$$

- Working it out

$$
X'y = \begin{bmatrix} 1 & 0 & 0 \\
                      0 & 1 & 1 \\
                    \end{bmatrix}
      \begin{bmatrix} y_1 \\
                      y_2  \\
                      y_3 \\
                      \end{bmatrix}
    = \begin{bmatrix} y_1  \\
                      y_2 + y_3 \\
                    \end{bmatrix}
$$

## Saturated Model Estimates

- $\hat{\beta}$ 

$$
[X'X]^{-1}X'y = \begin{bmatrix} 1 & 0 \\
                      0 & 0.5
                      \end{bmatrix} 
                \begin{bmatrix} y_1  \\
                      y_2 + y_3 \\
                    \end{bmatrix}
              = \begin{bmatrix} \frac{y_1}{1} & \frac{y_2 + y_3}{2} \end{bmatrix}
$$

- In other words

$$
\hat{\beta}_k = \overline{y}_k = \sum_{\forall i : D_k = k} y_i
$$

# Conditional Expectations Properties

## The CEF is a random variable

$$
\begin{split}
E\{E[Y_i|X_i]\} =& \int E[Y_i|X_i] g_x(u) du \\
  =& \int \left[  \int t f_y(t|X_i=u) dt \right] g_x(u) du \\
  =& \int \int t f_y(t|X_i=u)  g_x(u) du dt \\
  =& \int t \left[ \int  f_y(t|X_i=u)  g_x(u) du\right] dt \\
  =& \int t \left[ h_{xy}(u,t) du \right] dt  = \int t k_y(t) dt   = E[Y_i]
\end{split}
$$



## Decomposition Property

- Note from previous slide: 
    + Iterated expectations give us population parameters
    + We see this again, next slide
- For decomposition:
    - We can write $Y_i = E[Y_i|X_i] + \epsilon_i$
    - Because: $E[\epsilon_i|X_i] = 0$, $E[\epsilon h(X_i)] = 0$

## Additional math

- Condition on $X$

$$
\begin{split}
E[\epsilon|X_i] =& E\{Y_i-E[Y_i|X_i] |X_i \} \\
 =& E[Y_i|X_i] - E[Y_i|X_i]  \\
 =& 0
\end{split}
$$
- Conditioning on $X$ is same as conditiong on function of $X$

$$
E[\epsilon h(X_i)] = E \{E[\epsilon|X_i] h(X_i) \} = E[\epsilon|X_i]
h(X_i) = 0
$$


## CEF Prediction Property I

- Define $m(X_i)$ as a simple function of  $X_i$
- The CEF solves a minimum mean squared error (MMSE) prediction problem

$$
E[Y_i|X_i] = \underset{m(X_i)}{\operatorname{argmin}} \text{ }  E
\left[ \left( Y_i - m(X_i) \right)^2 \right] 
$$

## CEF Prediction Property II

- Note:

$$
\begin{aligned}
\left( Y_i - m(X_i) \right)^2 =& \left( (Y_i-E[Y_i|X_i]) + (E[Y_i|X_i]
  - m(X_i) \right)^2  \\
  =& (Y_i-E[Y_i|X_i])^2 + \\
   & 2 \{(E[Y_i|X_i]- m(X_i)) \times  (Y_i-E[Y_i|X_i]) \}\\
  &+ (E[Y_i|X_i]- m(X_i))^2
  \end{aligned}
$$

## CEF Prediction Property III

- Define $h(X_i) =  2 (E[Y_i|X_i]- m(X_i))$
$$
\begin{aligned}
\left( Y_i - m(X_i) \right)^2 =& (Y_i-E[Y_i|X_i])^2 \\
 &+ h(X_i) (Y_i-E[Y_i|X_i]) \\
 &+(E[Y_i|X_i]- m(X_i))^2
\end{aligned}
$$

- First term does not have $m$
- Second term is zero by 'rule' of CEF
- Third term, thus yields: $E[Y_i|X_i] = m(X_i)$

## ANOVA Theorem 

- Analysis of variance result

$$
V[Y_i] = V(E[Y_i|X_i])+E[V(Y_i|X_i)]
$$

- By CEF Decomposition and basic variance rule over a constant

$$
V[Y_i] = V(E[Y_i|X_i]) + V[\epsilon_i]
$$
- Explained sum of squares + unexplained sum of squares

# Frisch-Waugh-Lovell

## Some Initial Definitions

- Residual: $\hat{e} = y - \hat{y} = y - X\hat{\beta}$
- Prediction Maker: $P=X(X'X)^{-1}X'$
- Residual Maker: $M = I - X(X'X)^{-1}X'$
- Properties: 
    - Symmetric: $M'=M$
    - Idempotent: $MM = M'M = M$
    - Makes residual: $My = \hat{e}$
    - Regression of $X$ on $X$ gives perfect fit: $MX = 0$
    - ANOVA
    
## Residual and Prediction

- Makes residual

$$
\begin{aligned}
My &= y - X(X'X)^{-1}X'y \\
  &= y - X\hat{\beta} = y - \hat{y} \\
  &= \hat{e}
\end{aligned}
$$

- Prediction should be fairly clear

## ANOVA I

- Begin with a basic point

$$
\begin{aligned}
M+P &= I \\
(M+P)y &= Iy 
\end{aligned}
$$

## ANOVA II

- Use them to get our ANOVA result

$$
\begin{aligned}
\hat{e} + \hat{y} &= y \\
y'y &= (y'P + y'M)(Py + My) \\
 &= \hat{y}'\hat{y} + \hat{e}'\hat{e}
\end{aligned}
$$

- Total SS = Expl SS + Unexp SS

## A Simple Bivariate Model

- $X = [X_1 X_2]$
    - $X_1$ is a column of ones
    - $X_2$ is a single column random variable
- Define $M_1 = I - X_1(X_1'X_1)^{-1}X_1'$
    - It is a **residual maker** Matrix
    - It helps us understand an important regression point


## The basic theorem result I

- Observed and predicted (regression decomposition)
- Premultiply by M1

$$
\begin{aligned}
Y &= X_1 \hat{\beta}_1 + X_2 \hat{\beta}_2 + \hat{u} \\
M_1Y &= M_1X_1 \hat{\beta}_1 + M_1X_2 \hat{\beta}_2 + M_1\hat{u} \\
 M_1 Y &= 0 + M_1X_2 \hat{\beta}_2 + \hat{u}
\end{aligned}
$$

## The result II

- Now, premultiply by $X_2'$ and solve

$$
\begin{aligned}
 X_2' M_1 y &= X_2' M_1 X_2 \hat{\beta}_2 + 0 \\
 (X_2' M_1 X_2)^{-1} X_2' M_1 y &= (X_2' M_1 X_2)^{-1}X_2' M_1 X_2 \hat{\beta}_2  \\
 \hat{\beta}_2 &= (X_2' M_1 X_2)^{-1} X_2' M_1 y\\
   &= ((x_2^*) ' (x_2^*))^{-1} (x_2^*)' y
\end{aligned}
$$

## Meaning?

- We can get our estimate the usual way, or...
- We can regress $X_2$ on $X_1$ and obtain residuals $x_2^*$
- We then regress $y$ on $x_2^*$
- Notes: 
    - If $X_2$ and $X_1$ are not correlated, we do not need first regression
    - Thus, additional controls not always better
    - Multicollinearity shows up as small variation in $x_2^*$
    
## Aside I

- Regression of $X_2$ on $X_1$

$$
X_2 = X_1 \hat{\gamma} + x_2^*
$$

- Estimate

$$ 
\hat{\gamma} = (X_1'X_1)^{-1}X_1'X_2 
$$

## Aside II

- Residuals

$$
\begin{aligned}
x_2^* &= X_2 -  X_1 \hat{\gamma} \\
 &= X_2 - X_1 (X_1'X_1)^{-1}X_1'X_2 \\
 &= M_1 X_2
\end{aligned}
$$

- Since $M_1$ is idempotent..

$$
\begin{aligned}
X_2'M_1 X_2 &= X_2' M_1'M_1 X_2 \\
 &= (x_2^*) ' (x_2^*)
\end{aligned}
$$


# References 

## References {.allowframebreaks}







