---
title: "EKT 813 Project 1: NIDS Wave 2 Adult Health Dataset"
author: "Gustav, Nathan, Phakamani, Ruben"
format: html
editor: visual
execute:
  warning: false
  message: false
---

# Project 1

## Setup

```{r}
library(haven)
library(dplyr)
library(tidyr)
library(janitor)
library(lubridate)
library(tibble)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)
library(stringr)
```

## 1. Import data

We consider the Adult, Household Derived and Income Derived data sets.

```{r}
adult <- read_dta("~/GitHub/UPEconQuants/Data/Adult_W2_Anon_V4.0.0.dta") |>
  clean_names()

hhderived <- read_dta("~/GitHub/UPEconQuants/Data/hhderived_W2_Anon_V4.0.0.dta") |>
  clean_names()

indderived <- read_dta("~/GitHub/UPEconQuants/Data/indderived_W2_Anon_V4.0.0.dta") |>
  clean_names()
```

## 1.2 Clean Data

Codes are used to indicate observations in which the respondent refused to answer etc. We transform these coded values to missing observations, meaning they cannot interfere with the data accuracy. This is done at the start of the coding process to remove repetitive steps.

```{r}
codes <- c(-9, -8, -5, -3, -2,
           9999, 8888, 5555, 3333, 2222,
           99, 88, 55, 33, 22)

hhderived <- hhderived |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))

adult <- adult |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))

indderived <- indderived |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))
```

## 2. Build province and income inputs

What should be done to observations "other" contained in province data? Probably not in SA? So just say that they stay outside of SA? For the income data. If all the income components are missing, then the total remains missing. If at least 1 of the income components is 0 (with the rest being missing), the total is 0.

```{r}
#nids_missing_num <- c(-9, -8, -5, -3, -2)
#nids_missing_year <- c(9999, 8888, 5555, 3333, 2222)
#nids_missing_month_day <- c(99, 88, 55, 33, 22)


province_data <- hhderived |>
  select(w2_hhid, w2_prov2011) |>
  mutate(province = as_factor(w2_prov2011, levels = "labels")
  ) |>
  select(w2_hhid, province)


#income_components <- c(
#  "w2_fwag", "w2_cwag", "w2_swag", "w2_cheq", "w2_prof", "w2_extr",
#  "w2_bonu", "w2_othe", "w2_help", "w2_plot", "w2_opro", "w2_spen",
#  "w2_ppen", "w2_uif", "w2_comp", "w2_dis", "w2_chld", "w2_fost",
#  "w2_cdep", "w2_indi", "w2_inhe", "w2_rnt", "w2_retr", "w2_brid",
#  "w2_gift", "w2_loan", "w2_sale", "w2_remt"
#)

income_components <- indderived |>
  select(w2_fwag:w2_remt) |>
  names()



income_data <- indderived |>
  select(w2_hhid, pid, all_of(income_components)) |>
  mutate(
    income_nonmissing = rowSums(!is.na(across(all_of(income_components)))),
    income_rands = rowSums(across(all_of(income_components)), na.rm = TRUE),
    income_rands = if_else(income_nonmissing == 0, NA_real_, income_rands)
  ) |>
  select(w2_hhid, pid, income_rands)
```

## 3. Merge without dropping observations

Province data is merged using the HHID, which assumes that all respondents still live in that HH.

```{r}
#adult_merged <- adult |>
#  left_join(province_data, by = "w2_hhid") |>
#  left_join(income_data, by = c("w2_hhid", "pid"))

#stopifnot(nrow(adult_merged) == nrow(adult))

adult <- adult |>
  left_join(province_data, by = "w2_hhid") |>
  left_join(income_data, by = c("w2_hhid", "pid"))

stopifnot(nrow(adult) == nrow(adult))
```

## 4. Keep required variables, including interview outcome

```{r}
#symptom_vars <- c(
#  "w2_a_hl30fl", "w2_a_hl30fev", "w2_a_hl30pc", "w2_a_hl30cb", "w2_a_hl30tc",
#  "w2_a_hl30cp", "w2_a_hl30b", "w2_a_hl30h", "w2_a_hl30ba", "w2_a_hl30jp",
#  "w2_a_hl30v", "w2_a_hl30d", "w2_a_hl30w", "w2_a_hl30pua", "w2_a_hl30pla",
# "w2_a_hl30pu", "w2_a_hl30sa", "w2_a_hl30r", "w2_a_hl30sd", "w2_a_hl30c",
#  "w2_a_hl30wl", "w2_a_hl30ye", "w2_a_hl30ml", "w2_a_hl30i"
#)
symptoms <- adult |>
  select(w2_a_hl30fl:w2_a_hl30i) |>
  names()


#adl_vars <- c(
#  "w2_a_hldifdre", "w2_a_hldifbath", "w2_a_hldifeat", "w2_a_hldiftoil",
#  "w2_a_hldiftrn", "w2_a_hldifwrk", "w2_a_hldifmon", "w2_a_hldifstrs",
#  "w2_a_hldiflft", "w2_a_hldifwlk", "w2_a_hldifcook"
#)
adl_vars <- adult |>
  select(w2_a_hldifdre:w2_a_hldifcook) |>
  names()


#consult_vars <- c(
#  "w2_a_hlcontyp", "w2_a_hlconfee", "w2_a_hlconfee_v", "w2_a_hlconfee_p"
#)
consult_vars <- adult |>
  select(w2_a_hlcontyp:w2_a_hlconfee_p) |>
  names()


core_vars <- c(
  "w2_hhid", "pid", "w2_a_outcome",
  "w2_a_dob_y", "w2_a_dob_m", "w2_a_intrv_y", "w2_a_intrv_m", "w2_a_intrv_d",
  "w2_a_gen", "w2_a_popgrp", "w2_a_marstt", "w2_a_edschgrd",
  "w2_a_hlmedaid", "w2_a_hlcon",
  "province", "income_rands"
)


#adult_data_development <- adult_merged |>
#  select(any_of(c(core_vars, symptom_vars, adl_vars, consult_vars)))
adult_demographics <- adult |>
  select(any_of(c(core_vars, symptoms, adl_vars, consult_vars)))

```

## 5. Recode non-response values

This step is redundant. Test to make sure that the datasets are identical before and after recoding non-response values.

```{r}
#numeric_vars_to_clean <- setdiff(
#  names(adult_data_development |> select(where(is.numeric))),
#  c("w2_hhid", "pid")
#)

#adult_data_development <- adult_data_development |>
#  mutate(
#    across(
#      all_of(numeric_vars_to_clean),
#      ~ replace(.x, .x %in% nids_missing_num, NA_real_)
#    ),
#    across(
#      ~ replace(.x, .x %in% nids_missing_year, NA_real_)
#    ),
#    across(
#      c(w2_a_dob_m, w2_a_intrv_m, w2_a_intrv_d),
#      ~ replace(.x, .x %in% nids_missing_month_day, NA_real_)
#    )
#  )

#numeric_vars_to_clean <- setdiff(
#  names(adult_demographics |> 
#          select(where(is.numeric))),
#  c("w2_hhid", "pid")
#)

#adult_test <- adult_demographics |>
#  mutate(
#    across(
#      all_of(numeric_vars_to_clean),
#      ~ replace(.x, .x %in% codes, NA_real_)
#    ),
#    across(
#      c(w2_a_dob_y, w2_a_intrv_y),
#      ~ replace(.x, .x %in% codes, NA_real_)
#    ),
#    across(
#      c(w2_a_dob_m, w2_a_intrv_m, w2_a_intrv_d),
#      ~ replace(.x, .x %in% codes, NA_real_)
#    )
#  )

#identical(adult_demographics, adult_test)
```

## 6. Create age and health indicators

In ADL does 5 (able to but never do) a disability? I changed the name to reflect the dummies that are created.

```{r}
adult_demographics <- adult_demographics |>
  mutate(
    dob = make_date(w2_a_dob_y, w2_a_dob_m, 15),
    interview_date = make_date(w2_a_intrv_y, w2_a_intrv_m, w2_a_intrv_d),
    age_years = floor(time_length(interval(dob, interview_date), "years"))
  )

symptom_dummy <- adult_demographics |>
  select(all_of(symptoms)) |>
  mutate(
    across(
      everything(),
      ~ case_when(
        .x == 1 ~ 1L,
        .x == 2 ~ 0L,
        TRUE ~ NA_integer_
      )
    )
  )

adl_dummy <- adult_demographics |>
  select(all_of(adl_vars)) |>
  mutate(
    across(
      everything(),
      ~ case_when(
        .x %in% c(2, 3, 4) ~ 1L,
        .x %in% c(1, 5) ~ 0L,
        TRUE ~ NA_integer_
      )
    )
  )

 adult_demographics <- adult_demographics |>
  mutate(
    symptom_nonmissing = rowSums(!is.na(symptom_dummy)),
    symptom_yes_count = rowSums(symptom_dummy, na.rm = TRUE),
    ill30_any = case_when(
      symptom_nonmissing == 0 ~ NA_integer_,
      symptom_yes_count > 0 ~ 1L,
      TRUE ~ 0L
    ),
    adl_nonmissing = rowSums(!is.na(adl_dummy)),
    adl_yes_count = rowSums(adl_dummy, na.rm = TRUE),
    adl_any = case_when(
      adl_nonmissing == 0 ~ NA_integer_,
      adl_yes_count > 0 ~ 1L,
      TRUE ~ 0L
    )
  )
```

## 7. Create consultation variables

```{r}
adult_demographics <- adult_demographics |>
  mutate(
    annual_flag = case_when(
      w2_a_hlcon %in% c(1, 2, 3) ~ TRUE,
      w2_a_hlcon %in% c(4, 5, 6, 7, 8) ~ FALSE,
      TRUE ~ NA
    ),
    fee_charged_flag = case_when(
      w2_a_hlconfee == 1 ~ TRUE,
      w2_a_hlconfee == 2 ~ FALSE,
      TRUE ~ NA_real_
    ),
    consultation_fee_amount = case_when(
      annual_flag != TRUE ~ NA_real_,
      fee_charged_flag == FALSE ~ 0,
      fee_charged_flag == TRUE ~ w2_a_hlconfee_v,
      TRUE ~ NA_real_
    )
  )
```

## 8. Build the final adult_data_development dataset with cleaner variable names

```{r}
factor_vars <- c(
  "w2_a_outcome", "w2_a_gen", "w2_a_popgrp", "w2_a_marstt", "w2_a_edschgrd",
  "w2_a_hlmedaid", "w2_a_hlcon", "w2_a_hlcontyp", "w2_a_hlconfee", "w2_a_hlconfee_p"
)

adult_data_development <- adult_demographics |>
  mutate(
    across(all_of(factor_vars), ~ as_factor(.x, levels = "labels")),
    consultation_place = case_when(
      annual_flag == TRUE ~ as.character(w2_a_hlcontyp),
      TRUE ~ NA_character_
    ),
    consultation_fee_charged = case_when(
      annual_flag == TRUE ~ as.character(w2_a_hlconfee),
      TRUE ~ NA_character_
    ),
    consultation_paid_by = case_when(
      annual_flag == TRUE & fee_charged_flag == TRUE ~ as.character(w2_a_hlconfee_p),
      TRUE ~ NA_character_
    ),
    provider_type = case_when(
      consultation_place %in% c("Public hospital", "Public health clinic") ~ "Public",
      consultation_place %in% c("Private hospital", "Private clinic", "Private doctor") ~ "Private",
      consultation_place %in% c("Nurse or chemist", "Traditional healer") ~ "Other",
      TRUE ~ NA_character_
    ),
    interview_outcome = w2_a_outcome,
    gender = w2_a_gen,
    population_group = w2_a_popgrp,
    marital_status = w2_a_marstt,
    highest_grade_completed = w2_a_edschgrd,
    medical_aid = w2_a_hlmedaid,
    last_health_consultation = w2_a_hlcon,
    illness_last_30_days = factor(ill30_any, levels = c(0, 1), labels = c("No", "Yes")),
    difficulty_daily_living = factor(adl_any, levels = c(0, 1), labels = c("No", "Yes")),
    annual_flag = factor(
      annual_flag,
      levels = c(FALSE, TRUE),
      labels = c("No", "Yes")
    ),
    consultation_place = factor(consultation_place),
    consultation_fee_charged = factor(consultation_fee_charged),
    consultation_paid_by = factor(consultation_paid_by),
    provider_type = factor(provider_type, levels = c("Public", "Private", "Other"))
  ) |>
  select(
    household_id = w2_hhid,
    person_id = pid,
    interview_outcome,
    age_years,
    gender,
    population_group,
    marital_status,
    highest_grade_completed,
    province,
    monthly_income_rands = income_rands,
    medical_aid,
    illness_last_30_days,
    difficulty_daily_living,
    last_health_consultation,
    annual_flag,
    consultation_place,
    provider_type,
    consultation_fee_charged,
    consultation_fee_amount,
    consultation_paid_by
  )
```

## 9. Data development description and cleaner variable names

The final `adult_data_development` data set keeps the full merged adult sample and uses clean analysis-ready variable names. Variables are allowed to be missing where the questionnaire or linked files do not provide valid information.

```{r}
data_development_dictionary <- tibble(
  clean_name = names(adult_data_development),
  description = c(
    "Household identifier",
    "Person identifier",
    "Adult interview outcome",
    "Age in completed years at interview date",
    "Gender",
    "Population group",
    "Marital status",
    "Highest completed school grade",
    "Province from the household-derived file",
    "Total monthly individual income from indderived components",
    "Medical aid coverage",
    "Any illness symptom reported in the last 30 days",
    "Any reported difficulty with daily living activities",
    "Timing of most recent health consultation",
    "Whether respondent consulted about health in the last 12 months",
    "Place of most recent eligible consultation",
    "Grouped provider type: Public, Private, or Other",
    "Whether a consultation fee was charged",
    "Consultation fee amount in Rand",
    "Who paid the consultation fee"
  )
)

kable(
  data_development_dictionary,
  caption = "Table 1: Cleaner variable names and descriptions in adult_data_development",
  align = c("l", "l")
) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, width = "30em")
```

## 10. Row-retention check

```{r}
row_check <- tibble(
  stage = c("Adult raw file", "Final analytic dataset"),
  rows = c(nrow(adult), nrow(adult_data_development))
)

kable(
  row_check,
  caption = "Table 2: Row check showing that observations are retained in the master dataset"
) |>
  kable_styling(full_width = FALSE)
```

## 11. Variable description table

```{r}
var_info <- tibble(
  variable_name = c(
    "household_id", "person_id", "interview_outcome", "age_years", "gender",
    "population_group", "marital_status", "highest_grade_completed", "province",
    "monthly_income_rands", "medical_aid", "illness_last_30_days",
    "difficulty_daily_living", "last_health_consultation", "consulted_last_12_months",
    "consultation_place", "provider_type", "consultation_fee_charged",
    "consultation_fee_amount", "consultation_paid_by"
  ),
  definition = c(
    "Household identifier",
    "Person identifier",
    "Adult interview outcome",
    "Age in completed years at interview date",
    "Gender",
    "Population group",
    "Marital status",
    "Highest completed school grade",
    "Province from household-derived file",
    "Total monthly individual income from indderived components",
    "Medical aid coverage",
    "Any illness symptom in the last 30 days",
    "Any difficulty with daily living activities",
    "When respondent last consulted someone about health",
    "Consulted about health in the last 12 months",
    "Place of most recent eligible consultation",
    "Grouped provider type: Public, Private, or Other",
    "Whether a consultation fee was charged",
    "Consultation fee amount in Rand",
    "Who paid the consultation fee"
  )
)

obs_counts <- tibble(
  variable_name = names(adult_data_development),
  usable_observations = sapply(adult_data_development, function(x) sum(!is.na(x))),
  missing_observations = sapply(adult_data_development, function(x) sum(is.na(x)))
)

variable_table <- var_info |>
  left_join(obs_counts, by = "variable_name")

kable(
  variable_table,
  caption = "Table 3: Description of variables used in the analysis",
  align = c("l", "l", "r", "r")
) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, width = "26em")
```

## 12. Save cleaned dataset

```{r}
write_dta(adult_data_development, "Adult_W2_Project1_final_republished.dta")
write.csv(adult_data_development, "Adult_W2_Project1_final_republished.csv", row.names = FALSE)
```

## 13. Figure 1: Medical aid coverage

```{r}
build_age_group <- function(x) {
  case_when(
    x >= 15 & x <= 24 ~ "15-24",
    x >= 25 & x <= 34 ~ "25-34",
    x >= 35 & x <= 44 ~ "35-44",
    x >= 45 & x <= 54 ~ "45-54",
    x >= 55 & x <= 64 ~ "55-64",
    x >= 65 ~ "65+",
    TRUE ~ NA_character_
  )
}

med_summary <- adult_data_development |>
  mutate(
    age_group_plot = build_age_group(age_years),
    age_group_plot = factor(age_group_plot, levels = c("15-24", "25-34", "35-44", "45-54", "55-64", "65+")),
    medical_aid_yes = case_when(
      as.character(medical_aid) == "Yes" ~ 1,
      as.character(medical_aid) == "No" ~ 0,
      TRUE ~ NA_real_
    )
  ) |>
  filter(
    !is.na(age_group_plot),
    !is.na(population_group),
    !is.na(gender),
    !is.na(medical_aid_yes)
  ) |>
  group_by(age_group_plot, population_group, gender) |>
  summarise(
    medical_aid_coverage = mean(medical_aid_yes) * 100,
    n = n(),
    .groups = "drop"
  )

fig1_medical_aid <- med_summary |>
  ggplot(aes(x = age_group_plot, y = medical_aid_coverage, fill = population_group)) +
  geom_col(position = "dodge") +
  facet_wrap(~ gender, ncol = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Medical Aid Coverage by Age, Population Group and Gender",
    x = "Age group",
    y = "Medical aid coverage (%)",
    fill = "Population group",
    caption = "Source: NIDS Wave 2 adult data. Author's calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

fig1_medical_aid

ggsave("Figure1_Medical_Aid_Coverage.png", fig1_medical_aid, width = 12, height = 8, dpi = 300)
```

## 14. Figure 2: Public vs private vs other consultation

```{r}
consult_summary <- adult_data_development |>
  mutate(
    age_group_plot = build_age_group(age_years),
    age_group_plot = factor(age_group_plot, levels = c("15-24", "25-34", "35-44", "45-54", "55-64", "65+"))
  ) |>
  filter(
    as.character(annual_flag) == "Yes",
    !is.na(age_group_plot),
    !is.na(population_group),
    !is.na(gender),
    !is.na(provider_type)
  ) |>
  group_by(age_group_plot, population_group, gender, provider_type) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(age_group_plot, population_group, gender) |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup()


fig2_provider_type <- consult_summary |>
  ggplot(aes(x = age_group_plot, y = percentage, fill = provider_type)) +
  geom_col(position = position_dodge2(preserve = "single", width = 0.7)) +
  facet_grid(gender ~ population_group) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Type of Health Consultation by Age, Population Group and Gender",
    x = "Age group",
    y = "Share of consultations (%)",
    fill = "Provider type",
    caption = "Source: NIDS Wave 2 adult data. Author's calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

fig2_provider_type

ggsave("Figure2_Consultation_Type.png", fig2_provider_type, width = 14, height = 9, dpi = 300)
```

## 15. Final checks

```{r}
glimpse(adult_data_development)
summary(adult_data_development$age_years)
table(adult_data_development$interview_outcome, useNA = "ifany")
table(adult_data_development$province, useNA = "ifany")
table(adult_data_development$illness_last_30_days, useNA = "ifany")
table(adult_data_development$difficulty_daily_living, useNA = "ifany")
```

##Save the new dataset

```{r}
save(adult_data_development, file = "Adult_Project1.RData")
```


#Project 2

## 1. Incorporating Expenditure data 
```{r}
#Food Expenditure:
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, food_expend = w2_expf),
            by = "w2_hhid")

#Total Household Monthly Expenditure
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, total_expend = w2_expenditure),
            by = "w2_hhid")

#Non-food Expenditure:
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, non_food_expend = w2_expnf),
            by = "w2_hhid")

adult_data_development <- adult_data_development |>
  left_join(adult_demographics |>
              select())
```



