---
title: "EKT 813 Project 2: NIDS Wave 2 Adult Health Dataset"
author: "Gustav, Nathan, Phakamani, Ruben"
format: html
editor: visual
execute:
  warning: false
  message: false
knitr:
  opts_chunk:
    echo: true
    eval: true
    cache: true
    concordance: true
    autodep: true
    comment: "#>"
    dpi: 300
    fig.align: "center"
    out.width: 75%
---

# Project 1
To do: Clean province data approach. No need to create another data set.
## Setup

```{r}
library(haven)
library(dplyr)
library(tidyr)
library(janitor)
library(lubridate)
library(tibble)
library(ggplot2)
library(knitr)
library(kableExtra)
library(scales)
library(stringr)
```

## 1. Import data

We consider the Adult, Household Derived and Income Derived data sets.

```{r}
adult <- read_dta("~/GitHub/UPEconQuants/Data/Adult_W2_Anon_V4.0.0.dta") |>
  clean_names()

hhderived <- read_dta("~/GitHub/UPEconQuants/Data/hhderived_W2_Anon_V4.0.0.dta") |>
  clean_names()

indderived <- read_dta("~/GitHub/UPEconQuants/Data/indderived_W2_Anon_V4.0.0.dta") |>
  clean_names()
```

## 1.2 Clean Data

Codes are used to indicate observations in which the respondent refused to answer etc. We transform these coded values to missing observations, meaning they cannot interfere with the data accuracy. This is done at the start of the coding process to remove repetitive steps. For Project 2, include 44 and 77.

```{r}
codes <- c(-9, -8, -5, -3, -2,
           9999, 8888, 5555, 3333, 2222,
           99, 88, 55, 33, 22, 44, 77)

hhderived <- hhderived |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))

adult <- adult |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))

indderived <- indderived |>
  mutate(across(
    where(is.numeric),
    ~ replace(.x, .x %in% codes, NA_real_)
  ))
```

## 2. Build province and income inputs

What should be done to observations "other" contained in province data? Probably not in SA? So just say that they stay outside of SA? For the income data. If all the income components are missing, then the total remains missing. If at least 1 of the income components is 0 (with the rest being missing), the total is 0.

```{r}
province_data <- hhderived |>
  select(w2_hhid, w2_prov2011) |>
  mutate(province = as_factor(w2_prov2011, levels = "labels")
  ) |>
  select(w2_hhid, province)

income_components <- indderived |>
  select(w2_fwag:w2_remt) |>
  names()

income_data <- indderived |>
  select(w2_hhid, pid, all_of(income_components)) |>
  mutate(
    income_nonmissing = rowSums(!is.na(across(all_of(income_components)))),
    income_rands = rowSums(across(all_of(income_components)), na.rm = TRUE),
    income_rands = if_else(income_nonmissing == 0, NA_real_, income_rands)
  ) |>
  select(w2_hhid, pid, income_rands)
```

## 3. Merge without dropping observations

Province data is merged using the HHID, which assumes that all respondents still live in that HH.

```{r}
adult <- adult |>
  left_join(province_data, by = "w2_hhid") |>
  left_join(income_data, by = c("w2_hhid", "pid"))

stopifnot(nrow(adult) == nrow(adult))
```

## 4. Keep required variables, including interview outcome

```{r}
symptoms <- adult |>
  select(w2_a_hl30fl:w2_a_hl30i) |>
  names()

adl_vars <- adult |>
  select(w2_a_hldifdre:w2_a_hldifcook) |>
  names()

consult_vars <- adult |>
  select(w2_a_hlcontyp:w2_a_hlconfee_p) |>
  names()

core_vars <- c(
  "w2_hhid", "pid", "w2_a_outcome",
  "w2_a_dob_y", "w2_a_dob_m", "w2_a_intrv_y", "w2_a_intrv_m", "w2_a_intrv_d",
  "w2_a_gen", "w2_a_popgrp", "w2_a_marstt", "w2_a_edschgrd",
  "w2_a_hlmedaid", "w2_a_hlcon",
  "province", "income_rands"
)

adult_demographics <- adult |>
  select(any_of(c(core_vars, symptoms, adl_vars, consult_vars)))

```

## 5. Create age and health indicators

```{r}
adult_demographics <- adult_demographics |>
  mutate(
    dob = make_date(w2_a_dob_y, w2_a_dob_m, 15),
    interview_date = make_date(w2_a_intrv_y, w2_a_intrv_m, w2_a_intrv_d),
    age_years = floor(time_length(interval(dob, interview_date), "years"))
  )

symptom_dummy <- adult_demographics |>
  select(all_of(symptoms)) |>
  mutate(
    across(
      everything(),
      ~ case_when(
        .x == 1 ~ 1L,
        .x == 2 ~ 0L,
        TRUE ~ NA_integer_
      )
    )
  )

adl_dummy <- adult_demographics |>
  select(all_of(adl_vars)) |>
  mutate(
    across(
      everything(),
      ~ case_when(
        .x %in% c(2, 3, 4, 5) ~ 1L,
        .x %in% c(1) ~ 0L,
        TRUE ~ NA_integer_
      )
    )
  )

 adult_demographics <- adult_demographics |>
  mutate(
    symptom_nonmissing = rowSums(!is.na(symptom_dummy)),
    symptom_yes_count = rowSums(symptom_dummy, na.rm = TRUE),
    ill30_any = case_when(
      symptom_nonmissing == 0 ~ NA_integer_,
      symptom_yes_count > 0 ~ 1L,
      TRUE ~ 0L
    ),
    adl_nonmissing = rowSums(!is.na(adl_dummy)),
    adl_yes_count = rowSums(adl_dummy, na.rm = TRUE),
    adl_any = case_when(
      adl_nonmissing == 0 ~ NA_integer_,
      adl_yes_count > 0 ~ 1L,
      TRUE ~ 0L
    )
  )
```

## 6. Create consultation variables

```{r}
adult_demographics <- adult_demographics |>
  mutate(
    annual_flag = case_when(
      w2_a_hlcon %in% c(1, 2, 3) ~ TRUE,
      w2_a_hlcon %in% c(4, 5, 6, 7, 8) ~ FALSE,
      TRUE ~ NA
    ),
    fee_charged_flag = case_when(
      w2_a_hlconfee == 1 ~ TRUE,
      w2_a_hlconfee == 2 ~ FALSE,
      TRUE ~ NA_real_
    ),
    consultation_fee_amount = case_when(
      annual_flag != TRUE ~ NA_real_,
      fee_charged_flag == FALSE ~ 0,
      fee_charged_flag == TRUE ~ w2_a_hlconfee_v,
      TRUE ~ NA_real_
    )
  )
```

## 7. Build the final adult_data_development dataset with cleaner variable names

```{r}
factor_vars <- c(
  "w2_a_outcome", "w2_a_gen", "w2_a_popgrp", "w2_a_marstt", "w2_a_edschgrd",
  "w2_a_hlmedaid", "w2_a_hlcon", "w2_a_hlcontyp", "w2_a_hlconfee", "w2_a_hlconfee_p"
)

adult_data_development <- adult_demographics |>
  mutate(
    across(all_of(factor_vars), ~ as_factor(.x, levels = "labels")),
    consultation_place = case_when(
      annual_flag == TRUE ~ as.character(w2_a_hlcontyp),
      TRUE ~ NA_character_
    ),
    consultation_fee_charged = case_when(
      annual_flag == TRUE ~ as.character(w2_a_hlconfee),
      TRUE ~ NA_character_
    ),
    consultation_paid_by = case_when(
      annual_flag == TRUE & fee_charged_flag == TRUE ~ as.character(w2_a_hlconfee_p),
      TRUE ~ NA_character_
    ),
    provider_type = case_when(
      consultation_place %in% c("Public hospital", "Public health clinic") ~ "Public",
      consultation_place %in% c("Private hospital", "Private clinic", "Private doctor") ~ "Private",
      consultation_place %in% c("Nurse or chemist", "Traditional healer") ~ "Other",
      TRUE ~ NA_character_
    ),
    interview_outcome = w2_a_outcome,
    gender = w2_a_gen,
    population_group = w2_a_popgrp,
    marital_status = w2_a_marstt,
    highest_grade_completed = w2_a_edschgrd,
    medical_aid = w2_a_hlmedaid,
    last_health_consultation = w2_a_hlcon,
    illness_last_30_days = factor(ill30_any, levels = c(0, 1), labels = c("No", "Yes")),
    difficulty_daily_living = factor(adl_any, levels = c(0, 1), labels = c("No", "Yes")),
    annual_flag = factor(
      annual_flag,
      levels = c(FALSE, TRUE),
      labels = c("No", "Yes")
    ),
    consultation_place = factor(consultation_place),
    consultation_fee_charged = factor(consultation_fee_charged),
    consultation_paid_by = factor(consultation_paid_by),
    provider_type = factor(provider_type, levels = c("Public", "Private", "Other"))
  ) |>
  select(
    household_id = w2_hhid,
    person_id = pid,
    interview_outcome,
    age_years,
    gender,
    population_group,
    marital_status,
    highest_grade_completed,
    province,
    monthly_income_rands = income_rands,
    medical_aid,
    illness_last_30_days,
    difficulty_daily_living,
    last_health_consultation,
    annual_flag,
    consultation_place,
    provider_type,
    consultation_fee_charged,
    consultation_fee_amount,
    consultation_paid_by
  )
```

## 8. Row-retention check

```{r}
row_check <- tibble(
  stage = c("Adult raw file", "Final analytic dataset"),
  rows = c(nrow(adult), nrow(adult_data_development))
)

kable(
  row_check,
  caption = "Table 2: Row check showing that observations are retained in the master dataset"
) |>
  kable_styling(full_width = FALSE)
```

## 9. Save cleaned dataset

```{r}
write_dta(adult_data_development, "Adult_W2_Project1_final_republished.dta")
write.csv(adult_data_development, "Adult_W2_Project1_final_republished.csv", row.names = FALSE)
```

## 10. Figure 1: Medical aid coverage

```{r}
build_age_group <- function(x) {
  case_when(
    x >= 15 & x <= 24 ~ "15-24",
    x >= 25 & x <= 34 ~ "25-34",
    x >= 35 & x <= 44 ~ "35-44",
    x >= 45 & x <= 54 ~ "45-54",
    x >= 55 & x <= 64 ~ "55-64",
    x >= 65 ~ "65+",
    TRUE ~ NA_character_
  )
}

med_summary <- adult_data_development |>
  mutate(
    age_group_plot = build_age_group(age_years),
    age_group_plot = factor(age_group_plot, levels = c("15-24", "25-34", "35-44", "45-54", "55-64", "65+")),
    medical_aid_yes = case_when(
      as.character(medical_aid) == "Yes" ~ 1,
      as.character(medical_aid) == "No" ~ 0,
      TRUE ~ NA_real_
    )
  ) |>
  filter(
    !is.na(age_group_plot),
    !is.na(population_group),
    !is.na(gender),
    !is.na(medical_aid_yes)
  ) |>
  group_by(age_group_plot, population_group, gender) |>
  summarise(
    medical_aid_coverage = mean(medical_aid_yes) * 100,
    n = n(),
    .groups = "drop"
  )

fig1_medical_aid <- med_summary |>
  ggplot(aes(x = age_group_plot, y = medical_aid_coverage, fill = population_group)) +
  geom_col(position = "dodge") +
  facet_wrap(~ gender, ncol = 2) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Medical Aid Coverage by Age, Population Group and Gender",
    x = "Age group",
    y = "Medical aid coverage (%)",
    fill = "Population group",
    caption = "Source: NIDS Wave 2 adult data. Author's calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

fig1_medical_aid

ggsave("Figure1_Medical_Aid_Coverage.png", fig1_medical_aid, width = 12, height = 8, dpi = 300)
```

## 11. Figure 2: Public vs private vs other consultation

```{r}
consult_summary <- adult_data_development |>
  mutate(
    age_group_plot = build_age_group(age_years),
    age_group_plot = factor(age_group_plot, levels = c("15-24", "25-34", "35-44", "45-54", "55-64", "65+"))
  ) |>
  filter(
    as.character(annual_flag) == "Yes",
    !is.na(age_group_plot),
    !is.na(population_group),
    !is.na(gender),
    !is.na(provider_type)
  ) |>
  group_by(age_group_plot, population_group, gender, provider_type) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(age_group_plot, population_group, gender) |>
  mutate(percentage = n / sum(n) * 100) |>
  ungroup()


fig2_provider_type <- consult_summary |>
  ggplot(aes(x = age_group_plot, y = percentage, fill = provider_type)) +
  geom_col(position = position_dodge2(preserve = "single", width = 0.7)) +
  facet_grid(gender ~ population_group) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), limits = c(0, 100)) +
  labs(
    title = "Type of Health Consultation by Age, Population Group and Gender",
    x = "Age group",
    y = "Share of consultations (%)",
    fill = "Provider type",
    caption = "Source: NIDS Wave 2 adult data. Author's calculations."
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  )

fig2_provider_type

ggsave("Figure2_Consultation_Type.png", fig2_provider_type, width = 14, height = 9, dpi = 300)
```

## 12. Final checks

```{r}
glimpse(adult_data_development)
summary(adult_data_development$age_years)
table(adult_data_development$interview_outcome, useNA = "ifany")
table(adult_data_development$province, useNA = "ifany")
table(adult_data_development$illness_last_30_days, useNA = "ifany")
table(adult_data_development$difficulty_daily_living, useNA = "ifany")
```

##Save the new dataset

```{r}
save(adult_data_development, file = "Adult_Project1.RData")
```

#Project 2

## 1. Incorporating Expenditure data

```{r}
#Food Expenditure:
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, food_expend = w2_expf),
            by = "w2_hhid")


#Total Household Monthly Expenditure
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, total_expend = w2_expenditure),
            by = "w2_hhid")

#Non-food Expenditure:
adult_demographics <- adult_demographics |>
  left_join(hhderived |>
              select(w2_hhid, non_food_expend = w2_expnf),
            by = "w2_hhid")

#Include expenditure data by the PID
adult_data_development <- adult_data_development |>
  left_join(adult_demographics |>
              select(person_id = pid, food_expend, total_expend, non_food_expend),
            by = "person_id")
```


## 2. Identifying the household head
```{r}
#Create a dummy variable. 1 if PID is the HH decision maker, or 0.
#What to do with shared decision makers?

adult_demographics <- adult_demographics |>
  mutate(head_dummy = if_else(pid == adult$w2_a_decdpid, 1, 0))

adult_data_development <- adult_data_development |>
  left_join(adult_demographics |>
              select(person_id = pid, head_dummy),
            by = "person_id")
```


## 3. Self-assessed health
```{r}
#Respondents perceived health status: 
adult_demographics <- adult_demographics |>
  mutate(self_health = as_factor(adult$w2_a_hldes, levels = "labels"))

adult_data_development <- adult_data_development |>
  left_join(adult_demographics |>
              select(person_id = pid, self_health),
            by = "person_id")
```


## 4. Data Description Table
```{r}
var_info <- tibble(
  variable_name = c(
    "household_id", "person_id", "interview_outcome", "age_years", "gender",
    "population_group", "marital_status", "highest_grade_completed", "province",
    "monthly_income_rands", "medical_aid", "illness_last_30_days",
    "difficulty_daily_living", "last_health_consultation", "consulted_last_12_months",
    "consultation_place", "provider_type", "consultation_fee_charged",
    "consultation_fee_amount", "consultation_paid_by", "food_expend", "total_expend",
    "non_food_expend", "head_dummy", "self_health"
  ),
  definition = c(
    "Household identifier",
    "Person identifier",
    "Adult interview outcome",
    "Age in completed years at interview date",
    "Gender",
    "Population group",
    "Marital status",
    "Highest completed school grade",
    "Province from household-derived file",
    "Total monthly individual income from indderived components",
    "Medical aid coverage",
    "Any illness symptom in the last 30 days",
    "Any difficulty with daily living activities",
    "When respondent last consulted someone about health",
    "Consulted about health in the last 12 months",
    "Place of most recent eligible consultation",
    "Grouped provider type: Public, Private, or Other",
    "Whether a consultation fee was charged",
    "Consultation fee amount in Rand",
    "Who paid the consultation fee",
    "Monthly household expenditure on food",
    "Monthly household expenditure",
    "Monthly household non-food expenditure",
    "1 If the respondent is considered a head of the household",
    "Self-assessed health status"
  )
)

obs_counts <- tibble(
  variable_name = names(adult_data_development),
  usable_observations = sapply(adult_data_development, function(x) sum(!is.na(x))),
  missing_observations = sapply(adult_data_development, function(x) sum(is.na(x)))
)

variable_table <- var_info |>
  left_join(obs_counts, by = "variable_name")

kable(
  variable_table,
  caption = "Table 3: Description of variables used in the analysis",
  align = c("l", "l", "r", "r")
) |>
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover")) |>
  column_spec(1, bold = TRUE) |>
  column_spec(2, width = "26em")
```


## 5. Income and Expenditure Density Functions
Observations with a missing medical aid status have been excluded. 
```{r}
hhheads <- adult_data_development |> 
  filter(head_dummy == 1,
         !is.na(medical_aid),
         !is.na(self_health))

#Income Density Function:
ggplot(hhheads, aes(x = log(monthly_income_rands), fill = factor(medical_aid))) + 
  geom_density(alpha = 0.4) +
  labs(
    x = "Log of Monthly Income (Rands)",
    y = "Density",
    fill = "Medical Aid", 
    title = "Income distribution by Medical Aid Coverage of Household Heads"
  ) +
  scale_fill_manual(values = c("red", "darkgreen"),
                    labels = c("No medical aid", "Medical Aid")) +
  theme_classic()


#Expenditure Density Function: 
ggplot(hhheads, aes(x = log(total_expend), fill = factor(medical_aid))) + 
  geom_density(alpha = 0.4) +
  labs(
    x = "Monthly Expenditure (Rands)",
    y = "Density",
    fill = "Medical Aid", 
    title = "Expenditure distribution by Medical Aid Coverage of Household Heads"
  ) +
  scale_fill_manual(values = c("red", "darkgreen"),
                    labels = c("No medical aid", "Medical Aid")) +
  theme_classic()


#CDF of Medical aid coverage by income:
ggplot(hhheads, aes(x = monthly_income_rands, colour = medical_aid)) +
  stat_ecdf() +
  labs(
    x = "Monthly Income",
    y = "Cumulative Distribution Function",
    colour = "Medical Aid",
    title = "CDF of Medical Aid Coverage by Income"
  ) +
  scale_colour_manual(values = c("red", "darkgreen"),
                    labels = c("No medical aid", "Medical Aid"))+
  theme_classic()

```


## 6. Self-assessed Density Function
```{r}
#Self-assessed health status density function:
ggplot(hhheads, aes(x = self_health, fill = medical_aid)) + 
  geom_bar(position = "dodge") +
  labs(
    x = "Self-Assessed Health Status",
    y = "Count",
    fill = "Medical Aid", 
    title = "Self-Assessed Health Status by Medical Aid Coverage of Household Heads"
  ) +
  scale_fill_manual(values = c("red", "darkgreen"),
                    labels = c("No medical aid", "Medical Aid")) +
  theme_minimal()

#ggplot(hhheads, aes(x = self_health)) + 
#  geom_bar(fill = "steelblue") +
#  facet_wrap(~ medical_aid) +
#  labs(
#    x = "Self-Assessed Health Status",
#    y = "Count",
#    title = "Health Distribution by Medical Aid (Household Heads)"
#  ) +
#  theme_minimal()
```


## 7. Concentration curve for self-assessed health status
```{r}
heads <- adult_data_development |>
  filter(head_dummy ==1,
         !is.na(monthly_income_rands),
         !is.na(self_health))


heads <- heads |>
  mutate(self_health_num = case_when(
    self_health == "Excellent" ~ 1,
    self_health == "Very good" ~ 2,
    self_health == "Good" ~ 3,
    self_health == "Fair" ~ 4,
    self_health == "Poor" ~ 5
  ))

heads <- heads |>
  arrange(monthly_income_rands) |>
  mutate(
    rank = row_number(),
    cum_pop = rank / n()
  )

heads <- heads |>
  mutate(
    health_share = self_health_num / sum(self_health_num, na.rm = TRUE),
    cum_health = cumsum(health_share)
  )

ggplot(heads, aes(x = cum_pop, y = cum_health)) +
  geom_line(color = "orange", linewidth = 1) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
  labs(
    x = "Cumulative Share of Population (Ranked by Income)",
    y = "Cumulative Share of Health (Worse Health)",
    title = "Concentration Curve for Self-Assessed Health"
  ) +
  theme_minimal()
```




